---
title: "SISVAN 2019"
---

```{r}
#| label: setup
#| include: false

source(here::here("R/quarto-setup.R"))
```

<!-- badges: start -->
[![Project Status: Inactive – The project has reached a stable, usable state but is no longer being actively developed; support/maintenance will be provided as time allows.](https://www.repostatus.org/badges/latest/inactive.svg)](https://www.repostatus.org/#inactive)
[![License: MIT](https://img.shields.io/badge/license-MIT-green)](https://choosealicense.com/licenses/mit/)
<!-- badges: end -->

## Overview

This analysis focused on plotting data related to the food consumption of Brazilian children aged 2 to 4 years, as monitored by Brazil's Food and Nutrition Surveillance System ([SISVAN](https://sisaps.saude.gov.br/sisvan/)) in 2019.

## Setting things up

```{r}
#| eval: false
#| output: false

library(checkmate, quietly = TRUE)
library(colorspace, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(geobr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(googlesheets4, quietly = TRUE)
library(grDevices, quietly = TRUE)
library(here, quietly = TRUE)
library(raster, quietly = TRUE)
library(RColorBrewer, quietly = TRUE)
library(rutils, quietly = TRUE) # danielvartan/rutils (GitHub)
library(terra, quietly = TRUE)
library(tidyr, quietly = TRUE)
library(tidyterra, quietly = TRUE)
library(sf, quietly = TRUE)
library(spatstat.geom, quietly = TRUE)
library(stringdist, quietly = TRUE)
library(stringr, quietly = TRUE)
library(viridis, quietly = TRUE)
```

```{r}
get_data <- function(
    sheet,
    ss = "1nnDzHKufJGujXiuCFb6Hj14k0HIt6JPV0Y2wnqa2eAc",
    sheet_ignore = c("Documentation", "Codebook", "Validation", "Template")
    ) {
  checkmate::assert_string(sheet)
  checkmate::assert_string(ss)
  checkmate::assert_character(sheet_ignore)
  rutils:::assert_internet()

  ss <- googlesheets4::gs4_get(ss)
  sheets <- ss$sheets$name[!ss$sheets$name %in% sheet_ignore]
  checkmate::assert_subset(sheet, sheets)

  if (!sheet == "City validation") {
    cities <- googlesheets4::read_sheet(
      ss = ss, 
      sheet = "City validation", 
      col_names = TRUE,
      col_types = "c",
      na = c("", "NA"), 
      trim_ws = TRUE, 
      skip = 0
    ) |> 
      magrittr::extract2("value")
  }
  
  out <- googlesheets4::read_sheet(
      ss = ss, 
      sheet = sheet, 
      col_names = TRUE, 
      col_types = "c",
      na = c("", "NA"), 
      trim_ws = TRUE, 
      skip = 0
    )
  
  if (!sheet == "City validation") {
    out |> 
      dplyr::mutate(n = as.integer(n)) |>
      dplyr::filter(!n == 0)
  } else {
    out
  }
}
```

```{r}
assert_color <- function(color, null_ok = FALSE, na_ok = FALSE) {
  color_pattern <- "(?i)^#[a-f0-9]{3}$|^#[a-f0-9]{6}$|^transparent$"
  
  checkmate::assert_flag(null_ok)
  checkmate::assert_flag(na_ok)
  
  name <- deparse(substitute(color))
  
  if (is.null(color) && isFALSE(null_ok)) {
    cli::cli_abort(
      paste0(
        "{.strong {cli::col_red(name)}} cannot be {.strong NULL}."
      )
    )
  }
  
  if (!is.null(color)) {
    if (is.na(color) && isFALSE(na_ok)) {
      cli::cli_abort(
        paste0(
          "{.strong {cli::col_red(name)}} cannot be {.strong NA}."
        )
      )
    }
  }
  
  if (!is.null(color) && !is.na(color) &&
      !color %in% grDevices::colors() && 
      !checkmate::test_string(color, pattern = color_pattern)) {
    cli::cli_abort(
      paste0(
        "{.strong {cli::col_red(name)}} is not a valid color code. ",
        "It must contain a hexadecimal color code or one of the ",
        "values in {.strong {cli::col_blue('grDevices::color()')}}."
      )
    )
  }

  invisible(NULL)
}
```


```{r}
assert_color_options <- function(
    color_low = NULL, color_high = NULL, viridis = NULL
  ) {
  viridis_choices <- c(
    "magma", "A", "inferno", "B", "plasma", "C", "viridis", "D",
    "cividis", "E", "rocket", "F", "mako", "G", "turbo", "H"
  )
  
  color_pattern <- "(?i)^#[a-f0-9]{6}$"
  
  checkmate::assert_string(color_low, pattern = color_pattern, null.ok = TRUE)
  checkmate::assert_string(color_high, pattern = color_pattern, null.ok = TRUE)
  checkmate::assert_choice(viridis, viridis_choices, null.ok = TRUE)
  
  if (is.null(color_low) && !is.null(color_high) ||
      !is.null(color_low) && is.null(color_high)) {
    cli::cli_abort(
      paste0(
        "You must provide both ", 
        "{.strong {cli::col_blue('color_low')}} and ",
        "{.strong {cli::col_red('color_high')}} ",
        "arguments at the same time."
      )
    )
  } else if ((!is.null(color_low) | !is.null(color_high)) && 
             !is.null(viridis)) {
    cli::cli_abort(
      paste0(
        "You can't use both ", 
        "{.strong {cli::col_blue('color_low/color_high')}} and ",
        "{.strong {cli::col_red('viridis')}} ",
        "arguments at the same time."
      )
    )
  } else {
    invisible(NULL)
  }
}
```

```{r}
add_viridis <- function(
    plot, 
    option, 
    color_na = NULL, 
    binned = FALSE,
    breaks = ggplot2::waiver(),
    point = FALSE
  ) {
  option_choices <- c(
    "magma", "A", "inferno", "B", "plasma", "C", "viridis", "D",
    "cividis", "E", "rocket", "F", "mako", "G", "turbo", "H"
  )
  
  checkmate::assert_class(plot, "gg")
  checkmate::assert_choice(option, option_choices, null.ok = TRUE)
  assert_color(color_na, null_ok = TRUE)
  checkmate::assert_flag(binned)
  checkmate::assert_multi_class(breaks, c("waiver", "numeric"))
  checkmate::assert_flag(point)
  
  if (isTRUE(point)) {
    plot + 
      viridis::scale_color_viridis(
        option = option,
        na.value = color_na,
        breaks = breaks
      )
  } else if (isTRUE(binned)) {
    plot +
      ggplot2::scale_fill_binned(
        type = "viridis",
        na.value = color_na,
        breaks = breaks
      )
  } else {
    plot + 
      viridis::scale_fill_viridis(
        option = option,
        na.value = color_na,
        breaks = breaks
      )
  }
}
```

```{r}
add_fill <- function(
    plot, 
    color_low, 
    color_high, 
    color_na = NULL,
    binned = FALSE,
    breaks = ggplot2::waiver(),
    point = FALSE
) {
  checkmate::assert_class(plot, "gg")
  assert_color(color_low, null_ok = TRUE)
  assert_color(color_high, null_ok = TRUE)
  assert_color(color_na, null_ok = TRUE)
  checkmate::assert_flag(binned)
  checkmate::assert_multi_class(breaks, c("waiver", "numeric"))
  checkmate::assert_flag(point)
  
  if (isTRUE(point)) {
    plot + 
      ggplot2::scale_color_continuous(
        low = color_low, 
        high = color_high,
        na.value = color_na,
        breaks = breaks
      )
  } else if (isTRUE(binned)) {
    plot + 
      ggplot2::scale_fill_binned(
        type = "gradient",
        low = color_low,
        high = color_high,
        na.value = color_na,
        breaks = breaks
      )
  } else {
    plot + 
      ggplot2::scale_fill_gradient(
        low = color_low, 
        high = color_high,
        na.value = color_na,
        breaks = breaks
      )
  }
}
```

```{r}
add_theme <- function(plot, theme) {
  theme_choices <- c("minimal", "void")
  
  checkmate::assert_class(plot, "gg")
  checkmate::assert_choice(theme, theme_choices, null.ok = TRUE)
  
  if (theme == "minimal") {
    plot + ggplot2::theme_minimal()
  } else if (theme == "void") {
    plot + ggplot2::theme_void()
  } else {
    invisible(NULL)
  }
}
```

```{r}
plot_sf <- function(data, color_border = NA) {
  checkmate::assert_data_frame(data)
  checkmate::assert_subset(c("abbrev_state", "name_muni", "n"), names(data))
  assert_color(color_border, na_ok = TRUE)
  
  data |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(
      ggplot2::aes(fill = n),
      color = color_border
    )
}
```

```{r}
plot_raster <- function(
    data, 
    color_na = "white", 
    color_border = "black",
    breaks = ggplot2::waiver()
  ) {
  checkmate::assert_data_frame(data)
  checkmate::assert_subset(c("abbrev_state", "name_muni", "n"), names(data))
  assert_color(color_na, null_ok = TRUE)
  assert_color(color_border, na_ok = TRUE)
  checkmate::assert_multi_class(breaks, c("waiver", "numeric"))
  rutils:::assert_internet()
  
  res <- as.data.frame(do.call(
    "rbind", 
    lapply(sf::st_geometry(data), sf::st_bbox))
  )
  
  data_points <- data |> sf::st_centroid() |> rutils:::shush()
  
  data_points <- dplyr::tibble(
    lon = sf::st_coordinates(data_points)[, 1],
    lat = sf::st_coordinates(data_points)[, 2],
    n = data$n
  ) |>
    # tidyr::drop_na() |>
    dplyr::mutate(
      n = dplyr::if_else(n == 0, NA, n),
      order = rank(n, ties.method = "first")
    ) |>
    dplyr::arrange(order)
  
  ras_dom <- raster::raster(
    xmn = min(res[["xmin"]]),
    xmx = max(res[["xmax"]]),
    ymn = min(res[["ymin"]]),
    ymx = max(res[["ymax"]]),
    resolution = 0.25,
    crs = sf::st_crs(data)
  )
  
  sp::coordinates(data_points) <- ~ lon + lat
  data_raster <- 
    data_points |> 
    raster::rasterize(ras_dom, "n", update = TRUE) |>
    terra::rast()
  
  geobr_data_br <- 
    geobr::read_country(year = 2019, showProgress = FALSE) |>
    rutils:::shush()
  
  plot <- data |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = geobr_data_br,
      fill = "transparent", # "white"
      color = color_border
    ) + 
    tidyterra::geom_spatraster(
      data = data_raster
    ) +
    ggplot2::geom_sf(
      data = geobr_data_br,
      fill = "transparent",
      color = color_border
    )
  
  plot |> add_viridis(option = "plasma", "transparent", FALSE, breaks, FALSE)
}
```

```{r}
plot_point <- function(
    data, 
    color_na = "white", 
    color_border = "white",
    breaks = ggplot2::waiver(),
    alpha = 0.7
  ) {
  checkmate::assert_data_frame(data)
  checkmate::assert_subset(c("abbrev_state", "name_muni", "n"), names(data))
  assert_color(color_na, null_ok = TRUE)
  assert_color(color_border, na_ok = TRUE)
  checkmate::assert_multi_class(breaks, c("waiver", "numeric"))
  checkmate::assert_number(alpha, lower = 0, upper = 1)
  rutils:::assert_internet()
  
  data_points <- data |> sf::st_centroid() |> rutils:::shush()
  
  data_points <- dplyr::tibble(
    lon = sf::st_coordinates(data_points)[, 1],
    lat = sf::st_coordinates(data_points)[, 2],
    n = data$n,
    order = rank(n, ties.method = "first")
  ) |>
    tidyr::drop_na() |>
    dplyr::arrange(order)
  
  geobr_data_br <- 
    geobr::read_country(year = 2019, showProgress = FALSE) |>
    rutils:::shush()
  
  data |>
    ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = geobr_data_br,
      fill = color_na, # "transparent"
      color = color_border
    ) +
    ggplot2::geom_point(
      data = data_points,
      mapping = ggplot2::aes(
        x = lon, 
        y = lat, 
        size = n, 
        color = n
      ),
      alpha = alpha
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(), 
      size = ggplot2::guide_legend(),
    ) +
    ggplot2::scale_size_continuous(
      range = c(0, 5),
      breaks = breaks
    ) +
    ggplot2::theme(legend.key = ggplot2::element_blank())
}
```

```{r}
plot_brazil_city_map <- function(
    data, 
    color_low = NULL,
    color_high = NULL,
    color_na = "white",
    color_border = NA,
    viridis = NULL,
    binned = TRUE,
    breaks = NULL,
    zero_na = FALSE,
    point = FALSE,
    alpha = 0.7,
    density = FALSE,
    text_size = NULL,
    theme = NULL
  ) {
  viridis_choices <- c(
    "magma", "A", "inferno", "B", "plasma", "C", "viridis", "D",
    "cividis", "E", "rocket", "F", "mako", "G", "turbo", "H"
  )
  
  theme_choices <- c("minimal", "void")
  
  checkmate::assert_tibble(data)
  checkmate::assert_subset(c("abbrev_state", "name_muni", "n"), names(data))
  assert_color(color_low, null_ok = TRUE)
  assert_color(color_high, null_ok = TRUE)
  assert_color(color_na, null_ok = TRUE)
  assert_color(color_border, na_ok = TRUE)
  checkmate::assert_choice(viridis, viridis_choices, null.ok = TRUE)
  checkmate::assert_flag(binned)
  checkmate::assert_multi_class(breaks, c("waiver", "numeric"), null.ok = TRUE)
  checkmate::assert_flag(zero_na)
  checkmate::assert_flag(point)
  checkmate::assert_number(alpha, lower = 0, upper = 1)
  checkmate::assert_flag(density)
  checkmate::assert_number(text_size, null.ok = TRUE)
  checkmate::assert_choice(theme, theme_choices, null.ok = TRUE)
  assert_color_options(color_low, color_high, viridis)
  rutils:::assert_internet()
    
  geobr_data <-
      geobr::read_municipality(year = 2019, showProgress = FALSE) |>
      rutils:::shush()

  if (is.null(color_low) || is.null(color_high)) {
    colors <- RColorBrewer::brewer.pal(5,"YlOrRd")
    color_low <- dplyr::first(colors) # "#ffebd6"
    color_high <- dplyr::last(colors) # "#A90F06"
  }
  
  data <- 
    geobr_data |>
    dplyr::left_join(data, by = c("abbrev_state", "name_muni"))
  
  if (isTRUE(zero_na)) {
    data <- data |> dplyr::mutate(n = ifelse(is.na(n), 0, n))
  }
  
  if (is.null(breaks)) {
    # breaks <- c(0, 250, 500, 750, 1000)
    
    breaks <- seq(
      signif(max(data$n, na.rm = TRUE) / 5, 1), 
      max(data$n, na.rm = TRUE), 
      by = signif(max(data$n, na.rm = TRUE) / 5, 1)
    ) |> 
      signif(1)
  }
  
  if (isTRUE(density)) {
    # Not implemented.
    # plot <- data |> plot_raster(color_na, color_border, breaks)
  } else if (isTRUE(point)) {
    plot <- data |> plot_point(color_na, color_border, breaks, alpha)
  } else {
    plot <- data |> plot_sf(color_border)
  }
  
  plot <- 
    plot +
    ggplot2::labs(
      x = "Longitude", 
      y = "Latitude",
      color = "Total",
      fill = "Total",
      size = "Total"
    ) +
    ggplot2::theme(text = ggplot2::element_text(size = text_size))

  if (is.null(viridis)) {
    plot <- 
      plot |> 
      add_fill(color_low, color_high, color_na, binned, breaks, point)
  } else {
    plot <- plot |> add_viridis(viridis, color_na, binned, breaks, point)
  }

  if (!is.null(theme)) plot <- plot |> add_theme(theme)
  
  print(plot)
  invisible(plot)
}
```

<!-- Validate city values (SISVAN versus `geobr`) -->

```{r}
#| eval: false
#| include: false

geobr_data <- geobr::read_municipality(year = 2019, showProgress = FALSE)
```

```{r}
#| eval: false
#| include: false

# Create a 'Temp' sheet before using this function.

googlesheets4::range_write(
  ss = "1nnDzHKufJGujXiuCFb6Hj14k0HIt6JPV0Y2wnqa2eAc",
  dplyr::tibble(value = geobr_data$name_muni),
  sheet = "Temp",
  range = "A1",
  col_names = FALSE,
  reformat = FALSE
)
```

```{r}
#| eval: false
#| include: false

data <- get_data("City validation")
```

```{r}
#| eval: false
#| include: false

match <- stringdist::amatch(
  paste0(tolower(data$key_abbrev_state), tolower(data$key)), 
  paste0(geobr_data$abbrev_state, geobr_data$name_muni) |> 
    tolower() |>
    iconv(from = 'UTF-8', to = 'ASCII//TRANSLIT'), 
  maxDist = 1
)

paired_data <- 
  data |>
  dplyr::mutate(
    value_uf = geobr_data$abbrev_state[match],
    value = geobr_data$name_muni[match]
  )|>
  dplyr::select(key_uf, key, value_uf, value)
```

```{r}
#| eval: false
#| include: false

googlesheets4::range_write(
  ss = "1nnDzHKufJGujXiuCFb6Hj14k0HIt6JPV0Y2wnqa2eAc",
  paired_data,
  sheet = "City validation",
  range = "A2",
  col_names = FALSE,
  reformat = FALSE
)
```

```{r}
breaks <- c(250, 500, 750, 1000)
```

```{r}
#| eval: false
#| include: false

data_dir <- here::here("data")
```


## Habit of having at least the three main meals of the day

```{r}
data <- get_data("Habits - Meals")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "habits_meals.rds"))
}
```


```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

### Alternatives

```{r}
data |> plot_brazil_city_map(binned = FALSE)
```

```{r}
data |> plot_brazil_city_map(point = TRUE, breaks = breaks, alpha = 0.5)
```

```{r}
data |> plot_brazil_city_map(
  color_na = "gray95",
  breaks = breaks, 
  theme = "void"
)
```

```{r}
data |> plot_brazil_city_map(viridis = "plasma", binned = FALSE)
```

```{r}
#| eval: false
#| include: false

data |> plot_brazil_city_map(density = TRUE, viridis = "plasma")
```

## Habit of having meals while watching television

```{r}
data <- get_data("Habits - TV")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "habits_tv.rds"))
}
```

```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

## Consumption of beans

```{r}
data <- get_data("Consumption - Beans")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "consumption_beans.rds"))
}
```

```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

## Consumption of fruit

```{r} 
data <- get_data("Consumption - Fruits")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "consumption_fruits.rds"))
}
```

```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

## Consumption of vegetables and greens

```{r}
data <- get_data("Consumption - Vegetables")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "consumption_vegetables.rds"))
}
```

```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

## Consumption of ultraprocessed foods (UPF)

```{r}
data <- get_data("Consumption - UPFs")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "consumption_upfs.rds"))
}
```

```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

## Consumption of sugary drinks

```{r}
data <- get_data("Comsumption - Sugary drinks")
```

```{r}
#| eval: false
#| inclued: false

if (checkmate::test_tibble(data)) {
  saveRDS(data, file.path(data_dir, "consumption_sugary_drinks.rds"))
}
```

```{r}
data |> plot_brazil_city_map(breaks = breaks)
```

<!-- ## References {.unnumbered} -->

<!-- ::: {#refs} -->
<!-- ::: -->
